---
title: "Meeting 1"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(error=TRUE)
```

To start a notebook, select File, New File, R Notebook
To insert a "code block", press the "Insert" button above this window.

```{r}
2 + 2
```
Here R carries out the action of adding 2 + 2. The result is "returned", which in this case means just printing it to the screen

## Variables and basic data types

Assignment (setting a variable refer to a value)
```{r}
a <- 5
```
The action of assignment doesn't return any result. 

The variable remains in the environment, so you can use it again:
```{r}
a + 3
```
Other mathematical operators:

- Multiplication
```{r}
5 * 5
```
- Division
```{r}
10 / 5
```
- Powers
```{r}
10 ^ 2
```

Don't forget that if you change something in one chunk of R code you have to re-run every other block afterwards that refers to it (look at the options in the "Run" button above).

Assignment can also be done with "=":
```{r}
a = 99
```
But n.b. logical comparision "==" (i.e. a test that something "is the same as" something else):
```{r}
a == 1
```

```{r}
a == 99
```

**Careful not to mix up "=" and "=="!** This is a frequent source of unexpected behaviour (also known as "bugs") in programmes.

Numerals are one type of entity R knows about. 
Strings (or "characters") are another type of entity.

```{r}
lang <- "Arabic"
lang
```
Try this yourself 
```{r}
a + lang
```

Note `lang` is a variable name, so you write it directly into the console/your script. A string is marked with inverted commas, e.g. "English".
```{r}
lang
Michaels_native_language
```
```{r}
Michaels_native_language <- "English"
Michaels_native_language
```
Entering a string, like a numeral, returns itself
```{r}
"A string"
```

Numbers can be strings, but only in quotes
```{r}
a <- "99"
a
```
You still can't add numerals to characters.
```{r}
a + 1
```
You also can't use "+" with two strings. The mathematic operators are only defined for actions with numerals:
```{r}
lang + a
```

## Variable names

Legal variable names have the letters a-z (upper and lowercase are different), as well as `_` and `.`

```{r}
apple <- TRUE
Apple <- FALSE
```
These are two different variables
```{r}
apple
```
```{r}
Apple
```
```{r}
# The following won't work.
Michaels_native_language <- English
# The word English here has no quotes, so R understands it to be intended 
# as a variable name. But this variable doesn't exist (yet)
```
```{r}
Michaels_native_language <- "English"
Michaels_native_language
```
Meaningful variable names are good, but the variable name shouldn't be an actual value, bbecause this could be confusing:
```{r}
English <- "Russian"
Michaels_native_language <- English
Michaels_native_language
```
## Vectors

A *vector* is an ordered group of values of the same type (you might want to call this a "list", but techically, in R, a list is something else):

To make a vector you use a function *c*

Functions have the format e.g. *c()*, so a vector made of 1, 2, and 3 look like this:
```{r}
c(1, 2, 3)
```
A vector of strings is made the same way:
```{r}
c("Arabic", "Swedish")
```
The built-in variable *month.name* is also a vector of strings
```{r}
month.name
```
You can refer to items in a vector by their index number:
```{r}
# n.b. square brackets for index!
month.name[11] 
# i.e. the 11th item in the month.name vector
```
You can put variables into a vector:
```{r}
a <- 20
b <- 19
c(a, b, b, a)
```
This is different from e.g.
```{r}
c("a", "b", "b", "a")
```
You can't have mixed types in a vector:
```{r}
c(10, "a word")
```
Very very unfortunately, some early programmer decided the best thing to do here was not to raise an error, but to silently change the numeral into a string.

Use vectors as indexes to select part of a vector:
```{r}
summer <- c(12, 1, 2)
month.name[summer]
```
Shorthand for making sequential vectors (of numerals)
```{r}
1:10
```
Operations on vectors act on all members of the vector:
```{r}
my_seq <- 1:5
5 * my_seq
```
Operations with two vectors work item by item:
```{r}
my_seq + c(2,3,9,0,1)
```
Fails if the length doesn't match
```{r}
my_seq + c(2,3,9,0)
```
**Unless** one is an even multiple of the other, then is just repeats the shorter one to match:
```{r}
a <- 1:6
b <- c(0,1)
a + b
```

This is a horrible design decision made by somebody decades ago. It must have seemed like a good idea at the time.

## Variable names

Legal variable names

1. Start with a character a-z, A-Z (case sensitive!)
2. Contain any number of further characters, numerals, or "." or "_"

Some legal variable names:
```
a
a1
a.1
a_1
aa
my.seq.1
my_seq_1
mySeq1
```
## Matrix

Like a vector, but in two dimensions (still all the same type)

## Data frame

For tabular data, like in a spreadsheet

- a table with multiple rows and columns
- every column has a single type

```{r}
participant.code <- c("a", "b", "c", "d")
sex <- c("f", "m", "f", "f")
age <- c(24, 24, 28, 27)
my.exp <- data.frame(participant.code, sex, age)
my.exp
```
You can refer to the columns using "$":
```{r}
my.exp$age
```
You can assign new columns:
```{r}
my.exp$index.number <- 1:4
my.exp
```
